# Process{#s02}

```{r, include=F, warning=F, message=F}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots
library(ggnewscale) # ggnewscale
library(plot3D) # 3d plotting
library(rgl) # rgl plotting

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data

# models
library(brms) # bayesian modelling

# utilities
library(rvest) # web scraping

library(pak)
# load them
c("NeonTreeEvaluation", "cloud2trees", "LadderFuelsR", "lasR", "leafR") %>% 
# install and load
  purrr::map(function(x){
    # locations
    df <- dplyr::tibble(
      p = c("NeonTreeEvaluation", "cloud2trees", "LadderFuelsR", "lasR", "leafR")
      , l = c(
        "weecology/NeonTreeEvaluation_package"
        , "georgewoolsey/cloud2trees"
        , "olgaviedma/LadderFuelsR"
        , "r-lidar/lasR"
        , "DRAAlmeida/leafR"
      )
    )
    # install if needed
    if(!require(x, character.only = T)){
      pak::pkg_install(
        pkg = df %>% dplyr::filter(tolower(p)==tolower(x)) %>% dplyr::pull(l)
        , upgrade = T
      )
    }
    # load
    library(x, character.only = T)
  })



# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

In this [prior section](#s02) we reviewed the methodologies to process lidar data, extract a tree list, and estimate CBH.

Presently, we will outline the end-to-end process for accomplishing this task given some example lidar data.

## Get some lidar data

The USGS has a [lidar data explorer](https://apps.nationalmap.gov/lidar-explorer/#/). This data is non-standardized so it may be challenging to build any portable processes.

I found an example area of interest and navigated to the [file storage directory](https://rockyweb.usgs.gov/vdelivery/Datasets/Staged/Elevation/LPC/Projects/CO_Southwest_NRCS_2018_D18/CO_Southwest_NRCS_B3_2018/LAZ/)

We'll use the `rvest` package to download the las files programmatically.

```{r}
# url
url <- "https://rockyweb.usgs.gov/vdelivery/Datasets/Staged/Elevation/LPC/Projects/CO_Southwest_NRCS_2018_D18/CO_Southwest_NRCS_B3_2018/LAZ/"
# rvest read html
html <- rvest::read_html(x = url)
# get parent element
chrs <- html %>% rvest::html_elements("pre")
# unnest children elements
# chrs %>% rvest::html_elements("a")
las_files <- chrs[1] %>% 
  rvest::html_elements("a") %>% 
  rvest::html_text2() %>% 
  stringr::str_subset(".*\\.(laz|las)$")
```

download files

```{r}
# increase the download timeout
options(timeout = max(3600, getOption("timeout")))
# function to download
dl_fn <- function(x, my_url = url, dest = "E:/SJNF_lidar"){
    # create delivery dir if needed
    if(!file.exists(file.path(dest))){
      dir.create(file.path(dest), showWarnings = FALSE)
    }
    # parameters to download
    u <- paste0(my_url, x)
    d <- paste0(file.path(dest),"/", x)
    # download it...if needed
    if(!file.exists(d)){
      # print
      message("downloading file: ", x, "at ", Sys.time())
      # download it
      download.file(
        url = u, destfile = d
        , quiet = F
        , mode = "wb"
      )
    }
    return(d)
  }
# map over the las files to read
dl_fn_ans <- las_files %>% 
  .[1:20] %>%
  purrr::map(\(x) dl_fn(x)) %>% 
  unlist()
# dl_fn_ans
```

where in the world is this data?

```{r}
las_ctg <- lidR::readLAScatalog(dl_fn_ans)
mapview::mapview(las_ctg@data)
```

let's read one file

```{r}
las <- lidR::readLAS(dl_fn_ans[13])
las
```

plot a sample
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# ## if want to get current rgl parameters
# par3d()$zoom
# par3d()$FOV
# par3d()$userMatrix %>% c()
# par3d()$windowRect
## set up for printing rgl graphics
r3dDefaults <- rgl::r3dDefaults
m  <- structure(c(
  0.7963561,0.1201767,-0.5927685,0.0000000,-0.5896538,0.3724167,-0.7166689,0.0000000,0.1346301
  ,0.9202518,0.3674389,0.0000000,0.0000000,0.0000000,0.0000000,1.0000000
), .Dim = c(4L, 4L))
r3dDefaults$FOV <- 30
r3dDefaults$userMatrix <- m
r3dDefaults$zoom <- 0.55
r3dDefaults$windowRect <- c(0,23,1536,864)
rgl::setupKnitr(autoprint = TRUE)
```

```{r, rgl = TRUE}
las %>% 
  lidR::clip_roi(
    las %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_buffer(-400)
  ) %>% 
  lidR::plot(
    color = "Z", breaks = "quantile", bg = "white", legend = T
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
```

### `cloud2trees` that lidar data

extract trees from the lidar data using `cloud2trees` to create some example tree-level point cloud data to test the `LadderFuelsR` function we define below with

```{r}
if(F){
  cloud2trees_ans <- cloud2trees::cloud2trees(
    input_las_dir = dl_fn_ans[13]
    , output_dir = "../data"
    , estimate_tree_dbh = T
    , keep_intrmdt = T
  )
}else{
  cloud2trees_ans <- list(
    crowns_sf = sf::st_read("../data/point_cloud_processing_delivery/final_detected_crowns.gpkg", quiet=T)
    , treetops_sf = sf::st_read("../data/point_cloud_processing_delivery/final_detected_tree_tops.gpkg", quiet=T)
  )
}
```

clean it for `LadderFuelsR` to attach treeID to the point cloud

```{r}
# the lidR::merge_spatial requires only polygons so we need to rid the multipolygons
crowns_sf_poly <-
  # start with only polygons
  cloud2trees_ans$crowns_sf %>% 
  dplyr::filter(sf::st_geometry_type(.)=="POLYGON") %>%
  # union on cleaned multipolygons
  dplyr::bind_rows(
    cloud2trees_ans$crowns_sf %>% 
      dplyr::filter(sf::st_geometry_type(.)=="MULTIPOLYGON") %>%
      sf::st_cast(to = "POLYGON", do_split = T, warn = F) %>% 
      dplyr::mutate(axxx = sf::st_area(.)) %>% # axxx is so we don't overwrite a column
      dplyr::group_by(treeID) %>% 
      dplyr::filter(axxx == max(axxx)) %>% # keep the biggest crown polygon by treeID
      dplyr::ungroup() %>% 
      dplyr::select(-axxx)
  ) %>% 
  # generate a treeID index because it needs to be numeric
  dplyr::ungroup() %>% 
  dplyr::mutate(
    treeID_backup = treeID
    , treeID = dplyr::row_number()
  )
```

```{r, include=FALSE, eval=FALSE}
crowns_sf_poly %>% sf::st_geometry_type() %>% table()
cloud2trees_ans$crowns_sf %>% sf::st_geometry_type() %>% table()
```

read in normalized las files and filter one for testing

```{r}
# read in catalog
crowns_nlas_ctg <- lidR::readLAScatalog("../data/point_cloud_processing_temp/02_normalize/")
# filter for single tree point cloud
one_tree_sf <- crowns_sf_poly %>% 
  # get one of the taller trees
  dplyr::filter(tree_height_m >= quantile(crowns_sf_poly$tree_height_m, probs = 0.9)) %>% 
  dplyr::slice_sample(n=1)
```

now we'll attach the `treeID` column to the normalized las file and keep only the points that fall within a tree crown.

```{r, include=FALSE, eval=TRUE}
lidR::opt_progress(crowns_nlas_ctg) <- F
```

```{r}
# clip the point cloud
nlas_one_tree <- lidR::clip_roi(
    las = crowns_nlas_ctg
    , geometry = one_tree_sf
  ) %>% 
  sf::st_set_crs(sf::st_crs(crowns_sf_poly)) %>% 
  lidR::merge_spatial(
    source = crowns_sf_poly
    , attribute = "treeID"
  ) %>% 
  lidR::filter_poi(!is.na(treeID))
# what is this data?
nlas_one_tree@data %>% dplyr::glimpse()
```


## Define `LadderFuelsR` Processing Function

In this [prior section](#ladr_s) we identified the minimum steps needed to get CBH using the `LadderFuelsR` and `leafR` packages. Unfortunately, extracting CBH from the point cloud following these methods is performed one-by-one for individual trees. This does not seem like something that would work well if many, many trees.

There are a few possible paths forward to get CBH from point cloud data:
* Extract CBH from the point cloud for all trees using the `LadderFuelsR` [methodology](#ladr_s)
* Extract CBH from the point cloud for a sample of trees using the `LadderFuelsR` [methodology](#ladr_s) and build a model to estimate the rest
* Use the [TreeMap 2016 data](https://doi.org/10.2737/RDS-2021-0074) to model CBH using a regional model

Let's build a function to combine the `LadderFuelsR` steps for estimating CBH from the individual tree point cloud. Our function will take a point cloud as input and return a data frame.

```{r, include=FALSE, eval=FALSE}
# re-do the crs if needed
c_crs <- nlas_one_tree %>% sf::st_crs()
c_crs$wkt
c_crs$epsg
c_crs
sf::st_crs(nlas_one_tree) <- paste0("EPSG:",c_crs$epsg)
nlas_one_tree %>% sf::st_crs()
x_crs <- nlas_one_tree %>% sf::st_crs()
x_crs$wkt
x_crs$epsg
x_crs
x_crs == c_crs

# write it
fn_temp <- paste0(tempdir(), "/temp.las") 
#
  f_temp <- nlas_one_tree %>% 
    lidR::writeLAS(file = fn_temp)
# 
leafR::lad.voxels(normlas.file = f_temp, grain.size = 2) %>% 
  dplyr::glimpse()
```

```{r}
ladderfuelsr_cbh <- function(
  las
  , treeID = NA
  # leafR::lad.voxels
  , voxel_grain_size_m = 2
  # LadderFuelsR::get_gaps_fbhs
  , dist_btwn_bins_m = 1 # step
  , min_fuel_layer_ht_m = 1.5 # min_height
  , lad_perc_gap = 25 # perc_gap
  , lad_perc_base = 25 # perc_base
  # LadderFuelsR::nextfn
) {
  # check if string to las/laz file
  if(inherits(las, "character")){
    if(!stringr::str_ends(las, ".*\\.(laz|las)$")){
      stop("must pass a .las|.laz file path -OR- an object of class LAS to the `las` parameter")
    }
    # set the file path
    f <- normalizePath(las)
  }else if(inherits(las, "LAS")){
    # have to write the las to a tempfile
    fn <- paste0(tempdir(), "/temp.las") 
    # check if has a treeID
    if(
      (names(las@data) %>% stringr::str_detect("treeID") %>% max())==1
    ){
      n <- las@data$treeID %>% unique() %>% length()
      if(n>1 & is.na(treeID)){
        stop("the treeID column has more than one tree detected. set the `treeID` parameter")
      }else if(is.na(treeID)){
        # set the treeID
        treeID <- las@data$treeID %>% unique()
        # write it
        f <- las %>% 
          lidR::filter_poi(treeID == treeID) %>% 
          lidR::writeLAS(file = fn) 
      }else{
        # write it
        f <- las %>% 
          lidR::filter_poi(treeID == treeID) %>% 
          lidR::writeLAS(file = fn) 
      }
    }else{
      # write it
      f <- las %>% lidR::writeLAS(file = fn)
    }
  }else{
    stop("must pass a .las|.laz file path -OR- an object of class LAS to the `las` parameter")
  }
  #######################################
  ### Step 0 - `leafR` steps
  #######################################
    # 1) `leafR::lad.voxels()` - use normalized las file to create 
        # a data frame of the 3D voxels information (xyz) with Leaf Area Density values
    # 2) `leafR::lad.profile()` - calculate the lad profile from 
        # the input lad.voxels (step 1)
    # 3) ensure that the data frame returned from `leafR::lad.profile()` 
        # has a column named `treeID` which uniquely identifies individual trees. 
        # also, that column has to be the first column (bad practice by the authors)
    
    ## leafR::lad.voxels
    lad_voxels <- leafR::lad.voxels(normlas.file = f, grain.size = voxel_grain_size_m)
    ## leafR::lad.profile
    lad_profile <- leafR::lad.profile(lad_voxels, relative = F)
    ## add treeID column that is required by the package, though it's never stated
    lad_profile <- lad_profile %>% 
      dplyr::mutate(
        treeID = dplyr::coalesce(treeID, 1) %>% # if the treeID parameter is not set, fake 1
          factor()
      ) %>% 
      ## !!!!! not only does the treeID column have to exist...it has to be the first column
      dplyr::relocate(treeID)
    
  #######################################
  ### Step 1 - `LadderFuelsR::get_gaps_fbhs`
  #######################################
    ### this function is broken!!!!!!!!!!!!!!!!!!!!!!!!!!
    source("_broken_gap_fbh_from_LadderFuelsR.R") # gw fix here
    ## LadderFuelsR::get_gaps_fbhs
    gaps_fbhs <- 
      gw_get_gaps_fbhs(
      # LadderFuelsR::get_gaps_fbhs(
        LAD_profiles = lad_profile
        , step = dist_btwn_bins_m
        , min_height = min_fuel_layer_ht_m
        , perc_gap = lad_perc_gap
        , perc_base = lad_perc_base
        , verbose = F
      )
    # fix the columns that should be numeric
    gaps_fbhs <- gaps_fbhs %>% 
      dplyr::mutate(dplyr::across(
        !tidyselect::starts_with("treeID")
        , as.numeric
      ))    
    
  # return
    return(
      gaps_fbhs = gaps_fbhs
      , lad_profile = lad_profile
    )
}

# CALL IT
ladderfuelsr_cbh(
  las = nlas_one_tree
  # , treeID = unique(nlas_one_tree@data$treeID)
) %>% 
dplyr::glimpse()
```

```{r}
LadderFuelsR::get_plots_gap_fbh(
  LAD_profiles = lad_profile
  , gap_cbh_metrics = gaps_fbhs
  , min_height = 1.5
)
```

in RED are the GAPS and in GREEN the FBHs

can we re-create this?

```{r}
ggplot() + 
  geom_path(data = lad_profile, mapping = aes(x = lad, y = height)) +
  geom_point(data = lad_profile, mapping = aes(x = lad, y = height)) +
  # gaps data
  geom_hline(
    data = gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("gap") & !tidyselect::starts_with("gap_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "gaps")
    , linetype = "dotted"
    , lwd = 1.2
  ) + 
  # fbh data
  geom_hline(
    data = gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("cbh") & !tidyselect::starts_with("cbh_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "FBHs")
    , linetype = "dotdash"
    , lwd = 1.2
  ) +
  scale_color_manual(values = c("green4", "red"), name = "") +
  scale_y_continuous(breaks = scales::extended_breaks(10)) +
  theme_light() +
  theme(legend.position = "top")
```

yes. the `gap` and `cbh` columns contain the data ( but not the `gap_` and `cbh_` columns ;/ ).

but what are these columns?

* `cbh` - Height of the fuel layer base height (m)
* `gap` - Height of gap between fuel layers (m)

### Step 2 - `LadderFuelsR::calculate_gaps_perc`

this function calculates the percentile value of each height

```{r}
## LadderFuelsR::calculate_gaps_perc
#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! ERROR if treeID is not the first column
gaps_perc <- LadderFuelsR::calculate_gaps_perc(
  LAD_profiles = lad_profile %>% dplyr::select(-tree_index)
  , min_height=1.5
)
gaps_perc %>% dplyr::glimpse()
```

### Step 3 - `LadderFuelsR::get_distance`

calculates distances (and their heights) between fuel layers as the difference between consecutive gaps and fuel bases (the gap height always must be lower than the fuel base height).

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_distance()` calculates the distance between fuel layers as the height difference between each pair of consecutive gaps and FBHs (Figure 4a). In addition, when there are consecutive gaps, the distance is calculated as the difference between the minimum gap and the next FBH encountered (Figure 4b). Similarly, when there were consecutive FBHs, the distance was calculated as the difference between the minimum FBH and the previous gap encountered. When there are gaps above the last FBH, these values were removed for further analysis. (p.5)

```{r}
## LadderFuelsR::get_distance
metrics_distance <- LadderFuelsR::get_distance(
  gap_cbh_metrics = gaps_fbhs
  , gaps_perc = gaps_perc
  , step=1, min_height=1.5
)
metrics_distance %>% dplyr::glimpse()
```

### Step 4 - `LadderFuelsR::get_depths`

calculates fuels depth as the difference between gaps interleaved between fuel layers minus one step if the fuel depths are greater than one step.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_depths()` calculates the depth of each fuel layer as the difference between the gaps comprising the FBHs. Finally, to get the real layer depth, and only when layer depth is greater than the step, it subtracts the step value from the depth (Figure 5). There are special cases when there are no gaps between FBHs: (i) if the gap height is less than the minimum height of the FBHs, depth is calculated as the difference between the maximum height of the FBHs set and the height of that gap minus the step (Figure 6a); (ii) if the gap height is greater than the maximum height of FBHs, depth is the difference between the maximum gap height and the minimum height of the FBHs set minus the step (Figure 6b). (p.5)

```{r}
## LadderFuelsR::get_depths
metrics_depth <- LadderFuelsR::get_depths(
  LAD_profiles = lad_profile
  , distance_metrics = metrics_distance
  , step= 1, min_height= 1.5
)
metrics_depth %>% dplyr::glimpse()
```

### Step 5 - `LadderFuelsR::get_real_fbh`

reshapes fuel layers after removing distances equal to any number of height bin steps, keeping the first "base height" from those consecutive ones separated by such distance.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_real_fbh()` identifies the first FBH from consecutive FBHs or the first FBH from those separated by a distance equal to number of steps that can be skipped to reshape the fuel layers. For each distance value, it locates the next FBH value. If distance = num_step, the height of that FBH is propagated upwards. If distance > num_step, the height of that FBH is kept (Figure 7a,b). (p.5-6)

```{r}
## LadderFuelsR::get_real_fbh
real_fbh <- LadderFuelsR::get_real_fbh(
  depth_metrics = metrics_depth
  , step= 1, number_steps = 1, min_height=1.5
)
real_fbh %>% dplyr::glimpse()
```

### Step 6 - `LadderFuelsR::get_real_depths`

recalculates fuel layers depth after considering distances greater than the actual height bin step.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_real_depths()` calculates the cumulative heights of depth values when distance = num_step. It iterates over each distance value and if dist[i] > num_step, it keeps the corresponding depth value. However, if dist[i] = num_step, it sums all consecutive distances and corresponding depth values (Figure 7a,b). (p.6)

```{r}
## LadderFuelsR::get_real_depths
real_depth <- LadderFuelsR::get_real_depths(effective_fbh = real_fbh, step=1, min_height=1.5)
real_depth %>% dplyr::glimpse()
```

### Step 7 - `LadderFuelsR::get_effective_gap`

recalculates the distance between fuel layers after considering distances greater than any number of height bin steps.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_effective_gap()` calculates the effective distance between fuel layers based on the previously identified FBHs. It loops over all the FBHs and, at each iteration, it checks if the current value and the next value in FBHs are not equal. If they are not equal, it keeps the corresponding distance value, otherwise it removes it (Figure 8). (p.6)

```{r}
## LadderFuelsR::get_effective_gap
eff_gap <- LadderFuelsR::get_effective_gap(
  effective_depth = real_depth
  , number_steps = 1, min_height= 1.5
)
eff_gap %>% dplyr::glimpse()
```

### Step 8 - `LadderFuelsR::get_layers_lad`

calculates the percentage of Leaf Area Density (LAD) within each fuel layer (first output) and removes those fuel layers with LAD percentage less than a specified threshold (default 10 the depth of the remaining ones (second output).

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_layers_lad()` calculates LAD (%) within each fuel layer, which is defined by the height range between the FBH and its depth. First, it calculates the total LAD from the original profile. Next, it retrieves the beginning and end height bin of each fuel layer and calculates the percentage of total LAD that falls within that height range. Later, the fuel layers that had a LAD (%) less than a specified threshold (default 10%) were removed, recalculating the distances (Figure 9). (p.7)

```{r}
## LadderFuelsR::get_layers_lad
layers_lad_df <- LadderFuelsR::get_layers_lad(
  LAD_profiles = lad_profile
  , effective_distances = eff_gap
  , threshold=10,step = 1,min_height= 1.5
)
layers_lad_df %>% str()
```

idk why it is a list of 2 with the same data just the order of the `max_height` and `Hcbh1_Hdptf1` columns are switched. do you spot another difference??

looking through the befuddling README it looks like the authors only keep the second data frame in the list

```{r}
(layers_lad_df <- layers_lad_df[[2]])
```

is CBH in here? did we do it?

* treeID: tree ID with strings and numeric values
* treeID1: tree ID with only numeric values
* dptf: Depth of fuel layers (m) after considering distances greater than the actual height bin step
* effdist: Effective distance between consecutive fuel layers (m) after considering distances greater than any number of steps
* Hcbh: Base height of each fuel separated by a distance greater than the certain number of steps
* Hdptf: Height of the depth of fuel layers (m) after considering distances greater than the actual step
* Hdist: Height of the distance (> any number of steps) between consecutive fuel layers (m)
* Hcbh_Hdptf - Percentage of LAD values comprised in each effective fuel layer
* max_height - Maximum height of the tree profile
* nlayers - Number of effective fuel layers

is it possible to have multiple `Hcbh` values for one tree? is step 9 below even necessary at this point?

let's take the `LadderFuelsR::get_plots_effective()` function for a spin

```{r}
LadderFuelsR::get_plots_effective(
  LAD_profiles = lad_profile
  , effective_LAD = layers_lad_df
  , min_height = 1.5
)
```

maybe we'll try to recreate this plot at some point

```{r}
# wanna recreate this?
```

### Step 9 - ~~`LadderFuelsR::get_cbh_dist`~~ `LadderFuelsR::get_cbh_metrics`

`LadderFuelsR::get_cbh_dist` is described in the research article but does not exist in the package or README. Looks like `LadderFuelsR::get_cbh_metrics` is there though.

determines the CBH of a segmented tree using three criteria: maximum LAD percentage, maximum distance and the last distance.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_cbh_dist()` applies three criteria to define the CBH in a segmented tree: the fuel layer containing the maximum LAD (%) (Figure 10a), the fuel layer located at the highest distance (Figure 10b), and the fuel layer separated by the last effective distance (Figure 10c). In the case of maximum LAD (%), the output also gives the CBH from the second fuel layer when the first one has the maximum LAD (%) but its depth is smaller than the value indicated in the parameter ‘hdepth1_height’ (default 2m). (p.8)

```{r}
## LadderFuelsR::get_cbh_metrics
cbh_metrics <- LadderFuelsR::get_cbh_metrics(effective_LAD = layers_lad_df, min_height= 1.5)
cbh_metrics %>% dplyr::glimpse()
```

what are these?
* treeID: tree ID with strings and numeric values
* treeID1: tree ID with only numeric values
* dptf: Depth of fuel layers (m) after considering distances greater than the actual height bin step
* effdist: Effective distance between consecutive fuel layers (m) after considering distances greater than any number of steps
* Hcbh: Base height of each fuel separated by a distance greater than the certain number of steps
* Hdptf: Height of the depth of fuel layers (m) after considering distances greater than the actual step
* Hdist: Height of the distance (> any number of steps) between consecutive fuel layers (m)
* Hcbh_Hdptf - Percentage of LAD values comprised in each effective fuel layer
* maxlad_Hcbh - Height of the CBH of the segmented tree based on the maximum LAD percentage
* maxlad1_Hcbh - Height of the CBH from the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_Hcbh - Height of the CBH of the segmented tree based on the maximum distance found in its profile
* last_Hcbh - Height of the CBH of the segmented tree based on the last distance found in its profile
* maxlad_ - Values of distance and fuel depth and their corresponding heights at the maximum LAD percentage
* maxlad1_ - Values of distance and fuel depth and their corresponding heights for the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_ - Values of distance and fuel depth and their corresponding heights at the maximum distance
* last_ - Values of distance and fuel depth and their corresponding heights at the last distance
* nlayers - Number of effective fuel layers
* max_height - Maximum height of the tree profile

there are also some plotting functions

```{r}
# Generate plots for fuels LAD metrics
plots_cbh_maxlad <- LadderFuelsR::get_plots_cbh_LAD(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
plots_cbh_maxdist <- LadderFuelsR::get_plots_cbh_maxdist(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
plots_cbh_lastdist <- LadderFuelsR::get_plots_cbh_lastdist(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
# patchwork them
(plots_cbh_maxlad[[1]] + labs(title = "get_plots_cbh_LAD")) +
(plots_cbh_maxdist[[1]] + labs(title = "get_plots_cbh_maxdist")) +
(plots_cbh_lastdist[[1]] + labs(title = "get_plots_cbh_lastdist")) +
  patchwork::plot_layout(ncol = 2)
```

