# Process{#s02}

In this [prior section](#s02) we reviewed the methodologies to process lidar data, extract a tree list, and estimate CBH.

Presently, we will outline the end-to-end process for accomplishing this task given some example lidar data.

## Define Processing Functions

In this [prior section](#ladr_s) we identified the minimum steps needed to get CBH using the `LadderFuelsR` and `leafR` packages. Unfortunately, extracting CBH from the point cloud following these methods is performed one-by-one for individual trees. This does not seem like something that would work well if many, many trees.

There are a few possible paths forward to get CBH from point cloud data:
* Extract CBH from the point cloud for all trees using the `LadderFuelsR` [methodology](#ladr_s)
* Extract CBH from the point cloud for a sample of trees using the `LadderFuelsR` [methodology](#ladr_s) and build a model to estimate the rest
* Use the [TreeMap 2016 data](https://doi.org/10.2737/RDS-2021-0074) to model CBH using a regional model

Let's build a function to combine the `LadderFuelsR` steps for estimating CBH from the individual tree point cloud. Our function will take a point cloud as input and return a data frame.

```{r}
# las = "../data/point_cloud_processing_temp/02_normalize/RMNP_013_2018_normalize.laz"
# las <- crowns_nlas_temp
# class(las)
# inherits(las, "LAS")
# stringr::str_ends(las, ".*\\.(laz|las)$")
# (names(las@data) %>% stringr::str_detect("treeID") %>% max())==1

ladderfuelsr_cbh <- function(
  las, treeID = 1
  , voxel_grain_size_m = 2
) {
  # check if string to las/laz file
  if(inherits(las, "character")){
    if(!stringr::str_ends(las, ".*\\.(laz|las)$")){
      stop("must pass a .las or .laz file path to the `las` parameter")
    }
    # set the file path
    f <- normalizePath(las)
  }else if(inherits(las, "LAS")){
    # have to write the las to a tempfile
    fn <- paste0(tempdir(), "/temp.las") 
    # check if has a treeID
    if(
      (names(las@data) %>% stringr::str_detect("treeID") %>% max())==1
    ){
      n <- las@data$treeID %>% unique() %>% length()
      if(n>1){
        warning("the treeID column has more than one tree detected...results may not be what is expected")
      }
      # write it
      f <- las %>% 
        lidR::filter_poi(treeID == treeID) %>% 
        lidR::writeLAS(file = fn)
    }else{
      # write it
      f <- las %>% lidR::writeLAS(file = fn)
    }
  }else{
    stop("must pass a .las or .laz file path -OR- an object of class LAS to the `las` parameter")
  }
  #######################################
  ### Step 0 - `leafR` steps
  #######################################
    # 1) `leafR::lad.voxels()` - use normalized las file to create a data frame of the 3D voxels information (xyz) with Leaf Area Density values
    # 2) `leafR::lad.profile()` - calculate the lad profile from the input lad.voxels (step 1)
    # 3) ensure that the data frame returned from `leafR::lad.profile()` has a column named `treeID` which uniquely identifies individual trees. also, that column has to be the first column (bad practice by the authors)
    
    ## leafR::lad.voxels
    lad_voxels <- leafR::lad.voxels(normlas.file = f, grain.size = voxel_grain_size_m)
    ## leafR::lad.profile
    lad_profile <- leafR::lad.profile(lad_voxels, relative = F)
    ## add treeID column that is required by the package, though it's never stated
    lad_profile <- lad_profile %>% 
      dplyr::mutate(treeID = factor(treeID)) %>% 
      ## !!!!! not only does the treeID column have to exist...it has to be the first column
      dplyr::relocate(treeID)
    
  # return
    return(lad_profile)
}

ladderfuelsr_cbh(
  las = crowns_nlas_temp %>% lidR::filter_poi(tree_index == 22)
  , treeID = 22
) %>% 
dplyr::glimpse()
```

### Step 1 - `LadderFuelsR::get_gaps_fbhs`

calculates gaps and fuel layers base height (FBH) as the difference in percentiles between consecutive LAD values along the vertical tree profile (VTP). Negative differences are linked to gaps and positive differences to fuel base height.

**notice that this function is broken in the package...see [this issue](https://github.com/olgaviedma/LadderFuelsR/issues/2)**

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_gaps_fbhs()` identifies height bins with negative differences in LAD percentile for identifying gaps. When among consecutive height bins differences are negative, it selects the height bin with the lowest LAD (<= perc_gap [P25th]). Additionally, the function looks for height bins with LAD <= P25th, and when they are consecutive, it takes only the first and last values of the set being considered. For obtaining fuel layer base heights (FBHs), it identifies height bins with positive differences in LAD percentile. When such positive differences are consecutive, it selects the height bin with minimum LAD (> perc_base [P25th]). Moreover, it filters height bins with LAD > P25th, and when these are consecutive it takes only the first and last values of each set. Finally, to avoid any duplicated height bins in gaps and FBHs, it selects only the height bins identified as gaps that were not present in the FBH file, giving preference to the FBH height bins (p.3-5)

```{r}
### this function is broken!!!!!!!!!!!!!!!!!!!!!!!!!!
source("_broken_gap_fbh_from_LadderFuelsR.R") # gw fix here
## LadderFuelsR::get_gaps_fbhs
gaps_fbhs <- 
  gw_get_gaps_fbhs(
  # LadderFuelsR::get_gaps_fbhs(
    LAD_profiles = lad_profile
    , step=1
    , min_height=1.5
    , perc_gap= 25, perc_base= 25
    , verbose=TRUE
  )
gaps_fbhs %>% dplyr::glimpse()
# fix the columns that should be numeric
gaps_fbhs <- gaps_fbhs %>% 
  dplyr::mutate(dplyr::across(
    !tidyselect::starts_with("treeID")
    , as.numeric
  ))
```

let's take `LadderFuelsR::get_plots_gap_fbh()` for a spin which creates the plot in Figure 3 (plots of leaf area density profiles with fuel base heights in green and gaps >= step: Distance between bins in red).

>This function plots gaps and fuel layers base height (fbh) in the vertical tree profile (VTP).

```{r}
LadderFuelsR::get_plots_gap_fbh(
  LAD_profiles = lad_profile
  , gap_cbh_metrics = gaps_fbhs
  , min_height = 1.5
)
```

in RED are the GAPS and in GREEN the FBHs

can we re-create this?

```{r}
ggplot() + 
  geom_path(data = lad_profile, mapping = aes(x = lad, y = height)) +
  geom_point(data = lad_profile, mapping = aes(x = lad, y = height)) +
  # gaps data
  geom_hline(
    data = gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("gap") & !tidyselect::starts_with("gap_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "gaps")
    , linetype = "dotted"
    , lwd = 1.2
  ) + 
  # fbh data
  geom_hline(
    data = gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("cbh") & !tidyselect::starts_with("cbh_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "FBHs")
    , linetype = "dotdash"
    , lwd = 1.2
  ) +
  scale_color_manual(values = c("green4", "red"), name = "") +
  scale_y_continuous(breaks = scales::extended_breaks(10)) +
  theme_light() +
  theme(legend.position = "top")
```

yes. the `gap` and `cbh` columns contain the data ( but not the `gap_` and `cbh_` columns ;/ ).

but what are these columns?

* `cbh` - Height of the fuel layer base height (m)
* `gap` - Height of gap between fuel layers (m)

### Step 2 - `LadderFuelsR::calculate_gaps_perc`

this function calculates the percentile value of each height

```{r}
## LadderFuelsR::calculate_gaps_perc
#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! ERROR if treeID is not the first column
gaps_perc <- LadderFuelsR::calculate_gaps_perc(
  LAD_profiles = lad_profile %>% dplyr::select(-tree_index)
  , min_height=1.5
)
gaps_perc %>% dplyr::glimpse()
```

### Step 3 - `LadderFuelsR::get_distance`

calculates distances (and their heights) between fuel layers as the difference between consecutive gaps and fuel bases (the gap height always must be lower than the fuel base height).

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_distance()` calculates the distance between fuel layers as the height difference between each pair of consecutive gaps and FBHs (Figure 4a). In addition, when there are consecutive gaps, the distance is calculated as the difference between the minimum gap and the next FBH encountered (Figure 4b). Similarly, when there were consecutive FBHs, the distance was calculated as the difference between the minimum FBH and the previous gap encountered. When there are gaps above the last FBH, these values were removed for further analysis. (p.5)

```{r}
## LadderFuelsR::get_distance
metrics_distance <- LadderFuelsR::get_distance(
  gap_cbh_metrics = gaps_fbhs
  , gaps_perc = gaps_perc
  , step=1, min_height=1.5
)
metrics_distance %>% dplyr::glimpse()
```

### Step 4 - `LadderFuelsR::get_depths`

calculates fuels depth as the difference between gaps interleaved between fuel layers minus one step if the fuel depths are greater than one step.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_depths()` calculates the depth of each fuel layer as the difference between the gaps comprising the FBHs. Finally, to get the real layer depth, and only when layer depth is greater than the step, it subtracts the step value from the depth (Figure 5). There are special cases when there are no gaps between FBHs: (i) if the gap height is less than the minimum height of the FBHs, depth is calculated as the difference between the maximum height of the FBHs set and the height of that gap minus the step (Figure 6a); (ii) if the gap height is greater than the maximum height of FBHs, depth is the difference between the maximum gap height and the minimum height of the FBHs set minus the step (Figure 6b). (p.5)

```{r}
## LadderFuelsR::get_depths
metrics_depth <- LadderFuelsR::get_depths(
  LAD_profiles = lad_profile
  , distance_metrics = metrics_distance
  , step= 1, min_height= 1.5
)
metrics_depth %>% dplyr::glimpse()
```

### Step 5 - `LadderFuelsR::get_real_fbh`

reshapes fuel layers after removing distances equal to any number of height bin steps, keeping the first "base height" from those consecutive ones separated by such distance.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_real_fbh()` identifies the first FBH from consecutive FBHs or the first FBH from those separated by a distance equal to number of steps that can be skipped to reshape the fuel layers. For each distance value, it locates the next FBH value. If distance = num_step, the height of that FBH is propagated upwards. If distance > num_step, the height of that FBH is kept (Figure 7a,b). (p.5-6)

```{r}
## LadderFuelsR::get_real_fbh
real_fbh <- LadderFuelsR::get_real_fbh(
  depth_metrics = metrics_depth
  , step= 1, number_steps = 1, min_height=1.5
)
real_fbh %>% dplyr::glimpse()
```

### Step 6 - `LadderFuelsR::get_real_depths`

recalculates fuel layers depth after considering distances greater than the actual height bin step.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_real_depths()` calculates the cumulative heights of depth values when distance = num_step. It iterates over each distance value and if dist[i] > num_step, it keeps the corresponding depth value. However, if dist[i] = num_step, it sums all consecutive distances and corresponding depth values (Figure 7a,b). (p.6)

```{r}
## LadderFuelsR::get_real_depths
real_depth <- LadderFuelsR::get_real_depths(effective_fbh = real_fbh, step=1, min_height=1.5)
real_depth %>% dplyr::glimpse()
```

### Step 7 - `LadderFuelsR::get_effective_gap`

recalculates the distance between fuel layers after considering distances greater than any number of height bin steps.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_effective_gap()` calculates the effective distance between fuel layers based on the previously identified FBHs. It loops over all the FBHs and, at each iteration, it checks if the current value and the next value in FBHs are not equal. If they are not equal, it keeps the corresponding distance value, otherwise it removes it (Figure 8). (p.6)

```{r}
## LadderFuelsR::get_effective_gap
eff_gap <- LadderFuelsR::get_effective_gap(
  effective_depth = real_depth
  , number_steps = 1, min_height= 1.5
)
eff_gap %>% dplyr::glimpse()
```

### Step 8 - `LadderFuelsR::get_layers_lad`

calculates the percentage of Leaf Area Density (LAD) within each fuel layer (first output) and removes those fuel layers with LAD percentage less than a specified threshold (default 10 the depth of the remaining ones (second output).

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_layers_lad()` calculates LAD (%) within each fuel layer, which is defined by the height range between the FBH and its depth. First, it calculates the total LAD from the original profile. Next, it retrieves the beginning and end height bin of each fuel layer and calculates the percentage of total LAD that falls within that height range. Later, the fuel layers that had a LAD (%) less than a specified threshold (default 10%) were removed, recalculating the distances (Figure 9). (p.7)

```{r}
## LadderFuelsR::get_layers_lad
layers_lad_df <- LadderFuelsR::get_layers_lad(
  LAD_profiles = lad_profile
  , effective_distances = eff_gap
  , threshold=10,step = 1,min_height= 1.5
)
layers_lad_df %>% str()
```

idk why it is a list of 2 with the same data just the order of the `max_height` and `Hcbh1_Hdptf1` columns are switched. do you spot another difference??

looking through the befuddling README it looks like the authors only keep the second data frame in the list

```{r}
(layers_lad_df <- layers_lad_df[[2]])
```

is CBH in here? did we do it?

* treeID: tree ID with strings and numeric values
* treeID1: tree ID with only numeric values
* dptf: Depth of fuel layers (m) after considering distances greater than the actual height bin step
* effdist: Effective distance between consecutive fuel layers (m) after considering distances greater than any number of steps
* Hcbh: Base height of each fuel separated by a distance greater than the certain number of steps
* Hdptf: Height of the depth of fuel layers (m) after considering distances greater than the actual step
* Hdist: Height of the distance (> any number of steps) between consecutive fuel layers (m)
* Hcbh_Hdptf - Percentage of LAD values comprised in each effective fuel layer
* max_height - Maximum height of the tree profile
* nlayers - Number of effective fuel layers

is it possible to have multiple `Hcbh` values for one tree? is step 9 below even necessary at this point?

let's take the `LadderFuelsR::get_plots_effective()` function for a spin

```{r}
LadderFuelsR::get_plots_effective(
  LAD_profiles = lad_profile
  , effective_LAD = layers_lad_df
  , min_height = 1.5
)
```

maybe we'll try to recreate this plot at some point

```{r}
# wanna recreate this?
```

### Step 9 - ~~`LadderFuelsR::get_cbh_dist`~~ `LadderFuelsR::get_cbh_metrics`

`LadderFuelsR::get_cbh_dist` is described in the research article but does not exist in the package or README. Looks like `LadderFuelsR::get_cbh_metrics` is there though.

determines the CBH of a segmented tree using three criteria: maximum LAD percentage, maximum distance and the last distance.

([Viedma et al. 2024](https://doi.org/10.1111/2041-210X.14427)) note:

>Function `get_cbh_dist()` applies three criteria to define the CBH in a segmented tree: the fuel layer containing the maximum LAD (%) (Figure 10a), the fuel layer located at the highest distance (Figure 10b), and the fuel layer separated by the last effective distance (Figure 10c). In the case of maximum LAD (%), the output also gives the CBH from the second fuel layer when the first one has the maximum LAD (%) but its depth is smaller than the value indicated in the parameter ‘hdepth1_height’ (default 2m). (p.8)

```{r}
## LadderFuelsR::get_cbh_metrics
cbh_metrics <- LadderFuelsR::get_cbh_metrics(effective_LAD = layers_lad_df, min_height= 1.5)
cbh_metrics %>% dplyr::glimpse()
```

what are these?
* treeID: tree ID with strings and numeric values
* treeID1: tree ID with only numeric values
* dptf: Depth of fuel layers (m) after considering distances greater than the actual height bin step
* effdist: Effective distance between consecutive fuel layers (m) after considering distances greater than any number of steps
* Hcbh: Base height of each fuel separated by a distance greater than the certain number of steps
* Hdptf: Height of the depth of fuel layers (m) after considering distances greater than the actual step
* Hdist: Height of the distance (> any number of steps) between consecutive fuel layers (m)
* Hcbh_Hdptf - Percentage of LAD values comprised in each effective fuel layer
* maxlad_Hcbh - Height of the CBH of the segmented tree based on the maximum LAD percentage
* maxlad1_Hcbh - Height of the CBH from the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_Hcbh - Height of the CBH of the segmented tree based on the maximum distance found in its profile
* last_Hcbh - Height of the CBH of the segmented tree based on the last distance found in its profile
* maxlad_ - Values of distance and fuel depth and their corresponding heights at the maximum LAD percentage
* maxlad1_ - Values of distance and fuel depth and their corresponding heights for the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_ - Values of distance and fuel depth and their corresponding heights at the maximum distance
* last_ - Values of distance and fuel depth and their corresponding heights at the last distance
* nlayers - Number of effective fuel layers
* max_height - Maximum height of the tree profile

there are also some plotting functions

```{r}
# Generate plots for fuels LAD metrics
plots_cbh_maxlad <- LadderFuelsR::get_plots_cbh_LAD(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
plots_cbh_maxdist <- LadderFuelsR::get_plots_cbh_maxdist(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
plots_cbh_lastdist <- LadderFuelsR::get_plots_cbh_lastdist(
  LAD_profiles = lad_profile
  , cbh_metrics = cbh_metrics
  , min_height=1.5
)
# patchwork them
(plots_cbh_maxlad[[1]] + labs(title = "get_plots_cbh_LAD")) +
(plots_cbh_maxdist[[1]] + labs(title = "get_plots_cbh_maxdist")) +
(plots_cbh_lastdist[[1]] + labs(title = "get_plots_cbh_lastdist")) +
  patchwork::plot_layout(ncol = 2)
```

