# Height to Maximum Diameter Process{#s04}

In this section we'll use the benchmark data made available in the [`NeonTreeEvaluation`](https://github.com/weecology/NeonTreeEvaluation_package) data set ([Weinstein et al. 2021](https://scholar.google.com/scholar?cluster=4986448711981898434&hl=en&as_sdt=0,6)) to test a process for extracting the height of the maximum canopy diameter (HMD) using point cloud data. Our process will use individual trees and tree crowns detected from the point cloud, find the maximum width of point within the individual tree point cloud, and extract the height of those points. In short, we will determine HMD can by analyzing the vertical profile of the tree crown and identifying the height at which the crown diameter is maximum. The ultimate goal is to incorporate this process in the [`cloud2trees`](https://github.com/georgewoolsey/cloud2trees) package (Woolsey and Tinkham, 2024).

An early version of this process was developed by [Andrew Sanchez Meador](https://github.com/bi0m3trics).

First, load the standard libraries

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots
library(ggnewscale) # ggnewscale
library(rgl) # rgl plotting

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # tha cloud2trees
library(NeonTreeEvaluation) # benchmark data

```

```{r, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("Esri.WorldImagery", "OpenStreetMap")
  )
# clean session
remove(list = ls())
gc()
```

## Example Lidar Data{#rmnp_ex}

Let's load an example lidar dataset from Weinstein et al. ([2021](https://scholar.google.com/scholar?cluster=4986448711981898434&hl=en&as_sdt=0,6)) in their `NeonTreeEvaluation` package.

We'll use data from a NEON site that we know has conifers: RMNP (Rocky Mountain National Park)

```{r}
# let's see some field trees data
NeonTreeEvaluation::field %>% 
  dplyr::filter(siteID == "RMNP") %>% 
  dplyr::count(taxonID) %>% 
  dplyr::arrange(desc(n))
```

let's pick a site with the lowest proportion of POTR5

```{r}
plots_temp <- NeonTreeEvaluation::field %>% 
  dplyr::filter(siteID == "RMNP") %>% 
  dplyr::group_by(plotID) %>% 
  dplyr::summarise(
    trees = dplyr::n()
    , conifers = sum(ifelse(taxonID=="POTR5", 0, 1))
  ) %>% 
  dplyr::mutate(pct_conifer = conifers/trees) %>% 
  dplyr::filter(trees>20) %>% 
  dplyr::arrange(desc(pct_conifer), desc(trees))
plots_temp
```

get the lidar data

```{r}
# get the laz file path
las_f_path_temp <- paste0(system.file(package = "NeonTreeEvaluation"),"/extdata/") %>% 
    list.files(recursive = T, pattern = ".*\\.(laz|las)$", full.names = T) %>% 
    unique() %>% 
    dplyr::as_tibble() %>%
    dplyr::rename(f_path = 1) %>% 
    dplyr::filter(
      stringr::str_detect(f_path,  as.character(plots_temp[1,]$plotID))
    ) %>% 
    .[1] %>% # pulls the plot with the highest conifer proportion
    dplyr::pull(f_path)
# check the data
lidR::readLAS(las_f_path_temp)
```

```{r, include=FALSE, eval=FALSE}
las_xxx <- lidR::readLAS(las_f_path_temp)
epsg_xxx <- lidR::epsg(las_xxx)
wkt_xxx <- lidR::wkt(las_xxx)
crs_xxx <- lidR::crs(las_xxx)
```

### `cloud2trees` that lidar data

extract trees from the lidar data using `cloud2trees`

we do not need to get DBH or CBH for this testing

```{r, message=FALSE, results=FALSE}
# save our output_dir
od <- tempdir()
# cloud2trees
cloud2trees_ans <- cloud2trees::cloud2trees(
  input_las_dir = las_f_path_temp
  , output_dir = od
  , keep_intrmdt = T
)
```

```{r, include=FALSE, eval=FALSE}
las_zzz <- lidR::readLAS(
  list.files(
    file.path(od, "point_cloud_processing_temp", "02_normalize")
    , pattern = ".las"
    , full.names = T
  )[1]
)
epsg_zzz <- lidR::epsg(las_zzz)
wkt_zzz <- lidR::wkt(las_zzz)
crs_zzz <- lidR::crs(las_zzz)
# equality
identical(epsg_xxx, epsg_zzz)
identical(wkt_xxx, wkt_zzz)
identical(crs_xxx, crs_zzz)
```

let's see what we got

```{r}
cloud2trees_ans$treetops_sf %>% dplyr::glimpse()
```

where is this?

```{r}
# where?
cloud2trees_ans$treetops_sf %>% 
  mapview::mapview(layer.name = "trees")
```

*zoom out if you can't see anything*

## Height to Max Crown Diameter

The height of the maximum crown diameter (HMD) can be determined by analyzing the vertical profile of the tree crown and identifying the height at which the crown diameter is maximum. This can be achieved using techniques like profile analysis or by fitting parametric models to the crown shape. We'll use a simplified technique developed by [Andrew Sanchez Meador](https://github.com/bi0m3trics)

### One Tree Example

First, let's use our normalized point cloud for one tree which was dropped off in the output directory from the `cloud2trees::cloud2trees()` function (see that `output_dir` parameter).

```{r}
(n_f_temp <- list.files(
  file.path(od, "point_cloud_processing_temp", "02_normalize")
  , pattern = "(.las|.laz)"
  , full.names = T
))
# read in the lascatalog
nlas <- lidR::readLAScatalog(n_f_temp)
# we can set the options for reading from the catalog up-front
lidR::opt_progress(nlas) <- F
lidR::opt_filter(nlas) <- "-drop_duplicates -drop_class 2 9 18" ## class 2 = ground; 9 = water; 18 = noise
lidR::opt_select(nlas) <- "xyz"
# is there a treeID column?
nlas@data %>% names() %>% stringr::str_detect("treeID") %>% max() %>% as.logical()
```

clip the point cloud to one tree crown polygon

```{r}
# get id of tallest tree
id <- cloud2trees_ans$crowns_sf %>% dplyr::arrange(desc(tree_height_m)) %>% dplyr::slice_head(n=1) %>% dplyr::pull(treeID)
# keep only the one tree polygon
one_tree_sf <- cloud2trees_ans$crowns_sf %>% dplyr::filter(treeID==id)
# flter the point cloud and add the treeID
nlas_one_tree <- lidR::clip_roi(las = nlas, geometry = one_tree_sf) %>%
  lidR::add_attribute(x = id, name = "treeID")
```

plot the individual tree point cloud

```{r, include=FALSE, eval=FALSE}
# make a matrix to represent the cbh
x_temp <- seq(
    min(nlas_one_tree@data$X)
    , max(nlas_one_tree@data$X)
    , length.out = 2
  )
y_temp <- seq(
    min(nlas_one_tree@data$Y)
    , max(nlas_one_tree@data$Y)
    , length.out = 2
  )
xy_temp <- expand.grid(x = x_temp, y = y_temp)
z_temp <- matrix(
  rep(
    ladderfuelsr_cbh_ans$cbh_metrics$last_Hcbh
    , nrow(xy_temp)
  )
  , nrow = length(x_temp), ncol = length(y_temp)
)
# plot it
plot3D::scatter3D(
  x = nlas_one_tree@data$X
  , y = nlas_one_tree@data$Y
  , z = nlas_one_tree@data$Z
  , colvar = nlas_one_tree@data$Z
  , cex = 0.3, pch = 19
  , colkey = T
  , phi = -6
  , col = harrypotter::hp(n=50, house = "gryffindor")
  , main ="CBH shown in black"
  , surf = list(
    x = x_temp
    , y = y_temp
    , z = z_temp
    , facets = NA
    , border = "black"
    , lwd = 2
  )
)
```

```{r}
# plot it
plot3D::scatter3D(
  x = nlas_one_tree@data$X
  , y = nlas_one_tree@data$Y
  , z = nlas_one_tree@data$Z
  , colvar = nlas_one_tree@data$Z
  , cex = 0.3, pch = 19
  , colkey = T
  , phi = -6
  , col = harrypotter::hp(n=50, house = "gryffindor")
)
```

Walk through an example to get the height to maximum crown diameter (HMD)

```{r}
# # Example convex hull usage from graphics package
# X <- matrix(stats::rnorm(2000), ncol = 2)
# plot(X, cex = 0.5)
# (hpts <- grDevices::chull(X))
# (hpts <- c(hpts, hpts[1]))
# lines(X[hpts, ])


z <- nlas_one_tree$Z
x <- nlas_one_tree$X
y <- nlas_one_tree$Y

(i = base::which.max(z))
(xcenter = x[i])
(ycenter = y[i])

(A = lidR:::polygon_area(x, y)) ## internal lidR function
(n = length(x))
(xbar = mean(x))
(ybar = mean(y))
(height_tot = z[i])
  
# Calculate the convex hull of the crown points
(hull_indices <- grDevices::chull(x, y))
(hull_x <- x[hull_indices])
(hull_y <- y[hull_indices])
  
# Calculate the distances from the center to the hull points
(distances <- sqrt((hull_x - xcenter)^2 + (hull_y - ycenter)^2))
  
# Find the maximum distance and its corresponding height
(max_distance_index <- which.max(distances))
(max_crown_diameter_height <- z[hull_indices[max_distance_index]])
```


```{r}

# Define the myTreeMetrics function
myTreeMetrics <- function(x, y, z, i, treeID) {
  i = which.max(z)
  xcenter = x[i]
  ycenter = y[i]
  A = lidR:::polygon_area(x, y)
  n = length(x)
  xbar = mean(x)
  ybar = mean(y)
  height_tot = z[i]
  
  # Calculate the convex hull of the crown points
  hull_indices <- chull(x, y)
  hull_x <- x[hull_indices]
  hull_y <- y[hull_indices]
  
  # Calculate the distances from the center to the hull points
  distances <- sqrt((hull_x - xcenter)^2 + (hull_y - ycenter)^2)
  
  # Find the maximum distance and its corresponding height
  max_distance_index <- which.max(distances)
  max_crown_diameter_height <- z[hull_indices[max_distance_index]]
  
  metrics = list(
    npts = n,
    x = xcenter,
    y = ycenter,
    x_mean = xbar,
    y_mean = ybar,
    area_sqm = A,
    ht_m = height_tot,
    max_crown_diameter_height = max_crown_diameter_height
  )
  
  return(metrics)
}

# Example usage
LASfile <- system.file("extdata", "Megaplot.laz", package="lidR")
las = readLAS(LASfile)
las = segment_trees(las, li2012(R = 3, speed_up = 5))
tmetrics <- crown_metrics(las, ~myTreeMetrics(X, Y, Z, Intensity, treeID), geom = "point")

#===============================================================================
# Extract point data for 10 random trees and plot them
#===============================================================================
random_tree_ids <- sample(unique(las@data$treeID), 10)
tree_points <- las@data %>% filter(treeID %in% random_tree_ids)

# Calculate distances from the center for each tree
tree_points <- tree_points %>%
  group_by(treeID) %>%
  mutate(
    xcenter = mean(X[which.max(Z)]),
    ycenter = mean(Y[which.max(Z)]),
    distance = sqrt((X - xcenter)^2 + (Y - ycenter)^2)
  ) %>%
  ungroup()

# Calculate max_crown_diameter_height for each tree
max_crown_diameter_heights <- tree_points %>%
  group_by(treeID) %>%
  summarize(max_crown_diameter_height = max(Z[which.max(distance)]))

# Create ggplot
ggplot(tree_points, aes(x = distance, y = Z, color = Z)) +
  geom_point() +
  geom_hline(data = max_crown_diameter_heights, 
             aes(yintercept = max_crown_diameter_height), linetype = "dashed", color = "red") +
  scale_color_gradient(low = "blue", high = "green") +
  labs(title = "Random Trees",
       x = "Distance from Center",
       y = "Height (Z)",
       color = "Height (Z)") +
  facet_wrap(~ treeID, ncol = 5) +
  coord_equal() +
  theme_minimal()
```


let's put the `trees_species()` function in to `cloud2trees`

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(list = ls())
gc()
```
