# Tree Biomass Process{#s05}

In this section we'll use the benchmark data made available in the [`NeonTreeEvaluation`](https://github.com/weecology/NeonTreeEvaluation_package) data set ([Weinstein et al. 2021](https://scholar.google.com/scholar?cluster=4986448711981898434&hl=en&as_sdt=0,6)) to test a process for estimating biomass in kg at the tree level. The ultimate goal is to incorporate this process in the [`cloud2trees`](https://github.com/georgewoolsey/cloud2trees) package (Woolsey and Tinkham, 2024).

We will test out three methods for attaching biomass in kg to a spatial tree list:

1. Estimate single tree biomass (kg) using allometric equations based on predictors such as tree diameter, height, and species
    + This methodology is well represented in the literature with allometries developed for nearly all tree species ([Chojnacky et al. 2014](https://scholar.google.com/scholar?cluster=13769102016905655949&hl=en&as_sdt=0,5); [Ter-Mikaelian & Korzukhin 1997](https://scholar.google.com/scholar?cluster=6148302766104288498&hl=en&as_sdt=0,5))
    + [Gonzalez‐Akre et al. (2022)](https://scholar.google.com/scholar?cluster=14167690887599848846&oi=gsb&hl=en&as_sdt=0,5) developed an R package to perform this exact task: [`allodb`](https://github.com/ropensci/allodb).
2. Use models developed to predict canopy fuel stratum at the stand level based on common stand descriptors (e.g. TPH and  basal area) and distribute the stand-level estimate (in kg per m^2^ or kg per m^3^) across the individual trees in the stand
    + [Cruz et al. (2003)](https://scholar.google.com/scholar?cluster=316241498622221569&oi=gsb&hl=en&as_sdt=0,5) developed models to predict canopy fuel stratum at the stand level for four coniferous forest types common in the western US: Douglas-fir, ponderosa pine, lodgepole pine, and mixed conifer.
    + Models for other forests types are currently lacking which limits the scope of this methodology
3. Use raster estimates of canopy fuel stratum such as the [LANDFIRE](https://landfire.gov/fuel) database developed jointly by the USDA Forest Service and USDOI Office of Wildland Fire (OWF) and distribute the fuel estimate across the individual trees that fall in a raster cell
    + There is an R package for accessing the LANDFIRE data [`rlandfire`](https://github.com/bcknr/rlandfire)

## Preliminaries

Like [Cruz et al. (2003)](https://scholar.google.com/scholar?cluster=316241498622221569&oi=gsb&hl=en&as_sdt=0,5), we'll use the term "crown" to describe aerial fuels at the tree level and the term "canopy" at the stand level.

Below, we provide an overview of some commonly used fuel complex characteristics: canopy fuel load (CFL) and canopy bulk density (CBD). In simple terms, imagine a box filled with leaves, fuel load represents the total weight of all the leaves in the box while bulk density represents how tightly those leaves are packed within the box.

**Canopy Fuel Load**

* Definition: The total amount of flammable material (like leaves, needles, branches) within the tree crowns in a given area.
* Focus: Primarily on the mass of the fuel.
* Units: Typically expressed in units of mass per unit area (e.g., kilograms per square meter, tons per acre).
* Scope: Fuel load focuses on the overall amount of fuel.
* Significance: Important for understanding the overall fire potential of an area. A higher fuel load generally means more potential fuel for a fire to consume.

**Canopy Bulk Density**

* Definition: The mass of flammable material per unit volume of the tree crown.
* Focus: Considers both the mass of fuel and the space it occupies.
* Units: Typically expressed in units of mass per unit volume (e.g., kilograms per cubic meter).
* Scope: Bulk density considers how the fuel load is distributed within the tree crown.
* Significance: Crucial for predicting how easily a fire can spread through the tree crowns. Higher bulk density can increase the likelihood of crown fires.

## Data Load

load the libraries

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots
library(ggnewscale) # ggnewscale
library(rgl) # rgl plotting

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # tha cloud2trees
library(NeonTreeEvaluation) # benchmark data
library(allodb) # allometric biomass (kg) estimates
```

```{r, include=F, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("Esri.WorldImagery", "OpenStreetMap")
  )
# clean session
remove(list = ls())
gc()
```

## Example Lidar Data

Let's load an example lidar dataset from Weinstein et al. ([2021](https://scholar.google.com/scholar?cluster=4986448711981898434&hl=en&as_sdt=0,6)) in their `NeonTreeEvaluation` package.

We'll use data from a NEON site that we know has conifers: RMNP (Rocky Mountain National Park)

```{r}
# get the laz file path
las_f_path_temp <- system.file(package = "NeonTreeEvaluation", "extdata") %>% 
    list.files(recursive = T, pattern = ".*\\.(laz|las)$", full.names = T) %>% 
    unique() %>% 
    dplyr::as_tibble() %>%
    dplyr::rename(f_path = 1) %>% 
    dplyr::filter(
      stringr::str_detect(f_path,  "RMNP")
    ) %>% 
    dplyr::slice(15) %>% 
    dplyr::pull(f_path)
```

check the data

```{r}
# check the data
lidR::readLAS(las_f_path_temp) %>% 
  lidR::plot(
    color = "Z", breaks = "quantile", bg = "white", legend = T
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
```

### `cloud2trees` that lidar data

extract trees from the lidar data using `cloud2trees`

we will estimate DBH, CBH, and forest type for this data

```{r, message=FALSE, results=FALSE}
# save our output_dir
od <- tempdir()
# cloud2trees
cloud2trees_ans <- cloud2trees::cloud2trees(
  input_las_dir = las_f_path_temp
  , output_dir = od
  , keep_intrmdt = T
  , estimate_tree_dbh = T
  , estimate_tree_type = T
  , estimate_tree_cbh = T, cbh_estimate_missing_cbh = T, cbh_tree_sample_n = 55
)
```

```{r, include = F, eval = F, results=F, message=F}
cloud2trees_ans$treetops_sf <- cloud2trees::trees_dbh(cloud2trees_ans$treetops_sf)
```

let's see what we got

```{r}
cloud2trees_ans$treetops_sf %>% dplyr::glimpse()
```

where is this?

```{r}
# where?
cloud2trees_ans$treetops_sf %>% 
  mapview::mapview(layer.name = "trees")
```

*zoom out if you can't see anything*

what does our CBH data look like?

```{r}
cloud2trees_ans$treetops_sf %>% 
  ggplot(mapping = aes(x = tree_height_m, y = tree_cbh_m, color = is_training_cbh)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(x = "height (m)", y = "CBH (m)", color = "is CBH\nfrom cloud") +
  theme_light()
  
```

what does our DBH data look like?

```{r}
cloud2trees_ans$treetops_sf %>% 
  ggplot(mapping = aes(x = tree_height_m, y = dbh_cm)) +
  geom_point(color = "navy") +
  labs(x = "height (m)", y = "DBH (cm)") +
  theme_light()
  
```

how about the forest type?

```{r}
cloud2trees_ans$treetops_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(forest_type_group)
```

## Method 1: single tree biomass (kg) using allometric equations

The first method we'll explore for attaching biomass in kg to a spatial tree list is estimating single tree biomass (kg) using allometric equations based on predictors such as tree diameter, height, and species.

This methodology is well represented in the literature with allometries developed for nearly all tree species ([Chojnacky et al. 2014](https://scholar.google.com/scholar?cluster=13769102016905655949&hl=en&as_sdt=0,5); [Ter-Mikaelian & Korzukhin 1997](https://scholar.google.com/scholar?cluster=6148302766104288498&hl=en&as_sdt=0,5)). [Gonzalez‐Akre et al. (2022)](https://scholar.google.com/scholar?cluster=14167690887599848846&oi=gsb&hl=en&as_sdt=0,5) developed an R package to perform this exact task: [`allodb`](https://github.com/ropensci/allodb).

### Define species based on forest type group

`cloud2trees` gives us USDA Forest Inventory and Analysis (FIA) forest type group codes when we set the parameter `estimate_tree_type = T`. However, the `allodb` package requires genus and species information to estimate tree biomass. We need to make or find a lookup table to map the FIA forest type group to the genus and species.

Alternatively, we can use the `allodb::new_equations()` function to modify the set of equations that will be used to estimate the biomass

#### Manual

we can manually set the genus and species to PICO

```{r}
cloud2trees_ans$treetops_sf$biomass_kg <- allodb::get_biomass(
  dbh = cloud2trees_ans$treetops_sf$dbh_cm
  , genus = rep("Pinus", times = nrow(cloud2trees_ans$treetops_sf))
  , species = rep("contorta", times = nrow(cloud2trees_ans$treetops_sf))
  , coords = 
    cloud2trees_ans$treetops_sf %>%
      sf::st_transform(crs = sf::st_crs(4326)) %>%
      sf::st_coordinates() %>% 
      dplyr::as_tibble()
)
```

what did we get back?

```{r}
cloud2trees_ans$treetops_sf$biomass_kg %>% 
  summary()
```

let's plot biomass versus tree height because height comes directly from the point cloud

```{r}
cloud2trees_ans$treetops_sf %>% 
  ggplot(mapping = aes(x = tree_height_m, y = biomass_kg, color = biomass_kg)) +
  geom_point() +
  harrypotter::scale_color_hp(option = "slytherin") +
  scale_x_continuous(breaks = scales::extended_breaks(n = 11)) +
  scale_y_continuous(breaks = scales::extended_breaks(n = 11)) +
  labs(x = "height (m)", y = "biomass (kg)") +
  theme_light() +
  theme(legend.position = "none")

```

note that there is variation in the predicted biomass at a given height that is introduced from the variation in location of the trees

#### `allodb::new_equations()` function

the `allodb::new_equations()` function might allow us to modify the set of equations that will be used to estimate the biomass. can we adjust the weights used by `allodb::get_biomass()` to allow for multiple species allometries to contribute to the estimation of biomass given that we have only FIA forest type which groups multiple species together?

it looks like the `allodb::est_params()` returns the allometric equation parameters to be used to estimate biomass for each tree and it's returned data is joined with the tree list data based on the columns `c("genus", "long", "lat")` if no `species` parameter is provided (if `species` is provided, it is added to the join column list)

```{r}
# just get one coordinate location
coords_temp <- cloud2trees_ans$treetops_sf %>%
  sf::st_transform(crs = sf::st_crs(4326)) %>%
  sf::st_coordinates() %>% 
  dplyr::as_tibble() %>% 
  dplyr::slice(1)
# what parameters are default?
def_params_temp <- allodb::est_params(
  genus = "Pinus",
  coords = coords_temp,
  species = NULL,
  new_eqtable = NULL, # default for get_biomass
  wna = 0.1, # default for get_biomass
  w95 = 500, # default for get_biomass
  nres = 1e4 # default for get_biomass
)
def_params_temp
```

let's use `allodb::new_equations()` to filter the equations to use for biomass estimation

```{r}
new_eq_temp <- allodb::new_equations(
  subset_taxa = c(
    "Pinus" # in case the species doesn't exist
    , "Pinus contorta"
  )
)
```

what did we get?

```{r}
new_eq_temp %>% 
  dplyr::select(c("equation_id", "equation_taxa", "equation_allometry")) %>% 
  dplyr::glimpse()
```

now we'll update the `new_eqtable` parameter using the return from `allodb::new_equations()`

```{r}
updt_params_temp <- allodb::est_params(
  genus = "Pinus",
  coords = coords_temp,
  species = NULL,
  new_eqtable = new_eq_temp, # updated
  wna = 0.1, # default for get_biomass
  w95 = 500, # default for get_biomass
  nres = 1e4 # default for get_biomass
)
# what is the update?
updt_params_temp
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's figure out a mapping between genus/species and FIA forest types

```{r}
fia_trees <- system.file(package = "cloud2trees", "extdata", "treemap") %>% 
  list.files(full.names = T) %>% 
  stringr::str_subset("tree_table.csv") %>% 
  readr::read_csv(show_col_types = F, progress = F) %>% 
  dplyr::rename_with(tolower)
```

what is in the FIA tree table from TreeMap?

```{r}
fia_trees %>% 
  dplyr::glimpse()
```

we could overlay the TreeMap raster with the FIA forest type raster and aggregate the tree species by forest type......but at that point it seems like we might as well utilize the TreeMap FIA data from [Riley et al. (2021)](https://doi.org/10.2737/RDS-2021-0074) to build a [softmax regression (i.e. multinomial logistic regression)](https://bookdown.org/content/3686/nominal-predicted-variable.html) to predict the nominal variable tree species using the predictors tree height and location extracted from the point cloud data. This option roughly follows the process most represented in the existing literature as detailed above. Note that to predict the nominal variable tree species we could also use a random forest classifier as in [Meng et al. (2024)](https://doi.org/10.3390/f15122110).

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Method 2: stand-level canopy fuels models

The first method we'll explore for attaching biomass in kg to a spatial tree list is by using models developed to predict canopy fuel stratum at the stand level based on common stand descriptors (e.g. TPH and  basal area) and distribute the stand-level estimate (in kg per m^2^ or kg per m^3^) across the individual trees in the stand. The primary challenge with this methodology is that, currently a very limited set of forests types are represented in the literature by these stand-level models. This information gap limits the scope of this methodology.

### Cruz et al. 2003 equations

[Cruz et al. (2003)](https://scholar.google.com/scholar?cluster=316241498622221569&oi=gsb&hl=en&as_sdt=0,5) developed models to predict canopy fuel stratum at the stand level for four coniferous forest types common in the western US: Douglas-fir, ponderosa pine, lodgepole pine, and mixed conifer.

Let's build a function to utilize those models for data at the stand level to calculate Canopy Bulk Density (CBD) in kilograms per cubic meter

```{r}
get_cruz_stand_kg_per_m3 <- function(forest_type_group_code, basal_area_m2_per_ha, trees_per_ha){
  # Cruz et al. (2003)
  # https://scholar.google.com/scholar?cluster=316241498622221569&oi=gsb&hl=en&as_sdt=0,5
  # Page 46, Table 4
  if(as.numeric(forest_type_group_code) == 200){
    #Douglas-Fir Group
    b0 = -7.380
    b1 = 0.479
    b2 = 0.625
  }else if(as.numeric(forest_type_group_code) == 220){
    #Ponderosa Pine Group
    b0 = -6.649
    b1 = 0.435
    b2 = 0.579
  }else if(as.numeric(forest_type_group_code) == 280){
    #Lodgepole Pine Group
    b0 = -7.852
    b1 = 0.349
    b2 = 0.711
  }else if(as.numeric(forest_type_group_code) %in% c(120,260,320) ){
    #Mixed Conifer Group
    b0 = -8.445
    b1 = 0.319
    b2 = 0.859
  }else{
    #No Cruz et al. formulas for these YET!!!
    b0 = as.numeric(NA)
    b1 = as.numeric(NA)
    b2 = as.numeric(NA)
  }
  #Apply Cruz et al. if species found
  if(!is.na(b0)){
    return( 
      exp(b0 + b1 * log(basal_area_m2_per_ha) + b2 * log(trees_per_ha))
    )
  }else{
    return(as.numeric(NA))
  }
}
```

test our Cruz et al. (2003) `get_cruz_stand_kg_per_m3()` function

```{r}
get_cruz_stand_kg_per_m3(
  forest_type_group_code = 220 # ponderosa pine
  , basal_area_m2_per_ha = 25
  , trees_per_ha = 600
)
```

let's make a fake stand ID and data frame to show how to use this

```{r}
# fake a stand
cloud2trees_ans$treetops_sf <- cloud2trees_ans$treetops_sf %>% 
  dplyr::mutate(
    # ~2/3 will be "2", ~1/3 will be "1"
    stand_id = runif(n = nrow(cloud2trees_ans$treetops_sf), min = 0, max = 3) %>% 
      ceiling() %>% 
      ifelse(.==3,2,.)
  )
# aggregate
fake_stand_temp <- 
  cloud2trees_ans$treetops_sf %>% 
  sf::st_drop_geometry() %>% 
  # aggregate to stand level
  dplyr::group_by(stand_id, forest_type_group_code) %>% 
  dplyr::summarise(
    trees = dplyr::n()
    , basal_area_m2 = sum(basal_area_m2, na.rm = T)
  ) %>% 
  # fake per ha calcs
  dplyr::ungroup() %>% 
  dplyr::mutate(
    basal_area_m2_per_ha = basal_area_m2/0.1
    , trees_per_ha = trees/0.1
  )
fake_stand_temp
```

apply our Cruz et al. (2003) `get_cruz_stand_kg_per_m3()` function to the data frame

```{r}
fake_stand_temp <- fake_stand_temp %>% 
  dplyr::ungroup() %>% 
  dplyr::rowwise() %>% # this is key
  dplyr::mutate(
    kg_per_m3 = get_cruz_stand_kg_per_m3(
      forest_type_group_code = forest_type_group_code
      , basal_area_m2_per_ha = basal_area_m2_per_ha
      , trees_per_ha = trees_per_ha
    )
  ) %>% 
  dplyr::ungroup()
fake_stand_temp
```

looks good

### Define a "stand" using raster

for our `cloud2trees` process, we'll define a stand in the point cloud detected tree list using the USFS Forest Type Groups of the Continental United States raster Wilson (2023) available online [here](https://di-usfsdata.img.arcgis.com/arcgis/rest/services/CONUS_forest_type_group_2018_masked_202105122120120/ImageServer) with a data overview [here](https://www.arcgis.com/home/item.html?id=10760c83b9e44923bd3c18efdaa7319d) which was used to attach the tree species group to individual trees detected from the point cloud data based on spatial location.

let's read in the raster which we downloaded with `cloud2trees::get_data()`

```{r}
foresttype <- terra::rast(
  system.file(package = "cloud2trees", "extdata", "foresttype", "foresttype.tif")
)
```

what is this?

```{r}
foresttype
```

we will only use the data structure of this raster and not the values because we already have the forest type group in our tree list since we set the `estimate_tree_type = T` in our call to `cloud2trees::cloud2trees()`

```{r}
plt_temp <- foresttype %>% 
  terra::crop(
    cloud2trees_ans$treetops_sf %>% 
      sf::st_buffer(50) %>% 
      terra::vect() %>% 
      terra::project(terra::crs(foresttype))
  ) %>% 
  as.data.frame(xy=T) %>%
  ggplot() + 
  geom_tile(mapping = aes(x=x,y=y), fill = NA, color = "gray33") +
  theme_void() +
  theme(axis.text = element_blank(), axis.title = element_blank())
plt_temp
```

we only need to get the cell id to use this to group our tree list into "stands"

```{r}
cloud2trees_ans$treetops_sf$stand_id <- 
  # use terra::extract to get the cell id
  terra::extract(
    x = foresttype
    , y = cloud2trees_ans$treetops_sf %>% 
      terra::vect() %>% 
      terra::project(terra::crs(foresttype))
    , cells = T # cell numbers are also returned
  ) %>% 
  dplyr::pull(cell)
```

let's look at this spatially to make sure it makes sense

```{r}
plt_temp +
  cloud2trees_ans$treetops_sf %>% 
  sf::st_transform(crs = terra::crs(foresttype)) %>%
  geom_sf(mapping = aes(color = as.factor(stand_id))) +
  labs(color = "cell id\nwe'll call stand")
```

the challenge now is that we need to account for cases when our point cloud scan only overlaps a portion of a raster cell to scale our per hectare values

```{r}
las_ctg_sf <- sf::st_read(
  file.path(od, "point_cloud_processing_delivery", "raw_las_ctg_info.gpkg")
  , quiet = T
)
```

how does this look with our tree list?

```{r}
plt_temp <- plt_temp +
  # point cloud boundary
  geom_sf(
    data = las_ctg_sf %>% 
      sf::st_union() %>%
      sf::st_transform(crs = terra::crs(foresttype))
    , fill = NA, color = "red", lwd = 2
  ) +
  # tree points
  geom_sf(
    data = cloud2trees_ans$treetops_sf %>% 
      sf::st_transform(crs = terra::crs(foresttype))
    , mapping = aes(color = as.factor(stand_id))
  ) +
  labs(color = "cell id\nwe'll call stand")
plt_temp
```

we need to get a data frame of the cell numbers with the area of the raster cells that overlap with the point cloud extent. if the extent fully covers the raster cell, the area should be 1. if the extent only partially covers the cell, only the portion of the cell covered by the extent should be included in the area (e.g. 0.32).



### Distribute stand CBD (kg/m3) to trees

```{r}

```

