# Appendix

In this section we'll provide some more detail and work through examples for select processes to extract individual tree components using aerial point cloud data

```{r, include=F, warning=F, message=F}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots
library(ggnewscale) # ggnewscale
library(plot3D) # 3d plotting
library(rgl) # rgl plotting

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data

library(pak)
# load them
c("NeonTreeEvaluation", "cloud2trees", "LadderFuelsR", "lasR", "leafR") %>% 
# install and load
  purrr::map(function(x){
    # locations
    df <- dplyr::tibble(
      p = c("NeonTreeEvaluation", "cloud2trees", "LadderFuelsR", "lasR", "leafR")
      , l = c(
        "weecology/NeonTreeEvaluation_package"
        , "georgewoolsey/cloud2trees"
        , "olgaviedma/LadderFuelsR"
        , "r-lidar/lasR"
        , "DRAAlmeida/leafR"
      )
    )
    # install if needed
    if(!require(x, character.only = T)){
      pak::pkg_install(
        pkg = df %>% dplyr::filter(tolower(p)==tolower(x)) %>% dplyr::pull(l)
        , upgrade = T
      )
    }
    # load
    library(x, character.only = T)
  })



# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
# clean session
remove(list = ls())
gc()
```

## Example lidar data{#data_ex}

we'll use some of our trees and height-normalized point cloud data generated in our [point cloud processing](#processing)

load the tree crown polygons

```{r}
# get the data from already run
crowns_sf <-
  list.files(
    "../data/point_cloud_processing_delivery"
    , pattern = "final_detected_crowns.*\\.gpkg$"
    , full.names = T
  ) %>% 
  normalizePath() %>% 
  purrr::map(\(x)
    sf::st_read(
      dsn = x
      , quiet = T
    )
  ) %>% 
  dplyr::bind_rows()
```

load the central 0.1 ha area of our study area

```{r}
aoi <-
  sf::st_read("../data/point_cloud_processing_delivery/raw_las_ctg_info.gpkg") %>% 
  sf::st_union() %>% 
  sf::st_centroid() %>% 
  sf::st_buffer(sqrt(1000/4), endCapStyle = "SQUARE") %>% 
  sf::st_transform(sf::st_crs(crowns_sf))
```

filter for trees in the aoi

```{r}
crowns_sf <- crowns_sf %>% 
  # join so we get the full crown
  dplyr::inner_join(
    sf::st_intersection(crowns_sf, aoi) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::distinct(treeID)
    , by = "treeID"
  ) %>% 
  # make treeID numeric for later
  dplyr::mutate(
    treeID_bu = treeID
    , treeID = treeID %>% as.factor %>% as.numeric()
  ) %>% 
  cloud2trees::simplify_multipolygon_crowns()
```

let's check out the CHM of the trees in the aoi

```{r}
chm_rast <- terra::rast("../data/point_cloud_processing_delivery/chm_0.25m.tif")
# crop and plot
plt_chm <-
  chm_rast %>% 
  terra::crop(
    crowns_sf %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_transform(terra::crs(chm_rast)) %>% 
      terra::vect()
  ) %>% 
  terra::mask(
    crowns_sf %>% 
      sf::st_transform(terra::crs(chm_rast)) %>% 
      terra::vect()
  ) %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(ggplot2::aes(x=x,y=y,fill = f), color = NA) +
  ggplot2::labs(fill="CHM (m)") +
  harrypotter::scale_fill_hp("gryffindor") +
  ggplot2::theme_void()
# plot
plt_chm
```

```{r, include=FALSE, eval=T}
ggplot2::ggsave("../data/ex_chm.jpeg", height = 5, width = 5, dpi = "print")
```

and the DTM of the aoi

```{r}
dtm_rast <- terra::rast("../data/point_cloud_processing_delivery/dtm_1m.tif")
# crop and plot
dtm_rast %>% 
  terra::crop(
    crowns_sf %>% 
      sf::st_union() %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_buffer(100) %>% 
      sf::st_transform(terra::crs(dtm_rast)) %>% 
      terra::vect()
  ) %>% 
  terra::focal(w=5,fun="median",na.policy="omit") %>% 
  terra::as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_tile(ggplot2::aes(x=x,y=y,fill = f), color = NA) +
  ggplot2::labs(fill="DTM (m)") +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::theme_void()
```

```{r, include=FALSE, eval=T}
ggplot2::ggsave("../data/ex_dtm.jpeg", height = 5, width = 5, dpi = "print")
```

now, we'll load in the point cloud data for our aoi

```{r}
las_ctg <- lidR::readLAScatalog("../data/point_cloud_processing_delivery/norm_las/")
lidR::opt_progress(las_ctg) <- F
las <- lidR::clip_roi(las_ctg, crowns_sf %>% sf::st_bbox() %>% sf::st_as_sfc() %>% sf::st_buffer(2))
```

plot a sample

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# ## if want to get current rgl parameters
# par3d()$zoom
# par3d()$FOV
# par3d()$userMatrix %>% c()
# par3d()$windowRect
## set up for printing rgl graphics
r3dDefaults <- rgl::r3dDefaults
m  <- structure(c(
  0.90228689,-0.03516655,0.42969972,0.00000000,0.42887092,0.17524548,-0.88620484,0.00000000,-0.04413813,0.98389661
  ,0.17320365,0.00000000,0.00000000,0.00000000,0.00000000,1.00000000
), .Dim = c(4L, 4L))
r3dDefaults$FOV <- 30
r3dDefaults$userMatrix <- m
r3dDefaults$zoom <- 0.8
r3dDefaults$windowRect <- c(0,23,1536,864)
rgl::setupKnitr(autoprint = TRUE)
```

```{r, rgl = TRUE}
las %>% 
  lidR::plot(
    color = "Z", bg = "white", legend = F
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
```

here is the CHM + tree crowns

```{r}
plt_chm +
  ggplot2::geom_sf(data = crowns_sf %>% sf::st_transform(terra::crs(chm_rast)), fill = NA, color = "gray44", lwd = 1.5)
```

```{r, include=FALSE, eval=T}
ggplot2::ggsave("../data/ex_chm_trees.jpeg", height = 5, width = 5, dpi = "print")
```

now we'll attach the `treeID` column to the normalized las file

```{r}
las <- cloud2trees::polygon_attribute_to_las(las = las, poly_df = crowns_sf, attribute = "treeID", force_crs = T)
```

let's look at our point cloud colored by `treeID`

```{r, rgl = TRUE}
las %>% 
  lidR::filter_poi(!is.na(treeID)) %>% 
  lidR::plot(color = "treeID", bg = "white", legend = F)
```

let's filter for a single tree

```{r}
# sf
one_tree_sf <- crowns_sf %>% dplyr::filter(tree_height_m == max(tree_height_m)) %>% dplyr::slice(1)
# las
one_tree_las <- las %>% lidR::filter_poi(treeID==one_tree_sf$treeID)
```

check our one tree point cloud

```{r, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
# ## if want to get current rgl parameters
# par3d()$zoom
# par3d()$FOV
# par3d()$userMatrix %>% c()
# par3d()$windowRect
## set up for printing rgl graphics
r3dDefaults <- rgl::r3dDefaults
m  <- structure(c(
  1.0000000,0.0000000,0.0000000,0.0000000,0.0000000,0.3420201,-0.9396926,0.0000000
  ,0.0000000,0.9396926,0.3420201,0.0000000,0.0000000,0.0000000,0.0000000,1.0000000
), .Dim = c(4L, 4L))
r3dDefaults$FOV <- 30
r3dDefaults$userMatrix <- m
r3dDefaults$zoom <- 1.05
r3dDefaults$windowRect <- c(0,23,1536,864)
rgl::setupKnitr(autoprint = TRUE)

# ```{r, rgl = TRUE}
one_tree_las %>% 
  lidR::plot(
    color = "Z", bg = "white", legend = T
   , pal = harrypotter::hp(n=50, house = "gryffindor")
  )
```

```{r}
# plot it
plot3D::scatter3D(
  x = one_tree_las@data$X
  , y = one_tree_las@data$Y
  , z = one_tree_las@data$Z
  , colvar = one_tree_las@data$Z
  , cex = 0.6, pch = 19
  , colkey = T
  , phi = 11
  , col = harrypotter::hp(n=50, house = "gryffindor")
  , main ="single tree point cloud"
)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(las_ctg, aoi)
gc()
```

## CBH Process{#cbh}

Let's go through the process to extract CBH from the point cloud using the same settings as in our [point cloud processing](#processing). CBH is extracted directly from the height normalized point cloud using the process outlined in [Viedma et al. (2024)](https://doi.org/10.1111/2041-210X.14427) and implemented via `cloud2trees::ladderfuelsr_cbh()`. See `cloud2trees::trees_cbh()` for a simple way to extract tree CBH from a height normalized point cloud based on a tree list with crown polygons.

```{r, results=F, message=F}
# CALL IT
ladderfuelsr_cbh_ans <- cloud2trees::ladderfuelsr_cbh(
  las = one_tree_las
  , min_vhp_n = 3
  , voxel_grain_size_m = 1
  , dist_btwn_bins_m = 1
  , min_fuel_layer_ht_m = 1
  , lad_pct_gap = 25
  , lad_pct_base = 25
  , num_jump_steps = 1
  , min_lad_pct = 10
  , frst_layer_min_ht_m = 1
)
```

let's see what we got

```{r}
ladderfuelsr_cbh_ans %>% names()
ladderfuelsr_cbh_ans$lad_profile %>% dplyr::glimpse()
```

Create our own plot of the gaps and fuel layers base height in the vertical tree leaf area density (LAD) profile. [Almeida et al. (2019)](https://doi.org/10.3390/rs11010092) describe the process for generating leaf area density profiles and make the process accessible via the `leafR` [package](https://github.com/DRAAlmeida/leafR). In `cloud2trees`, methods to generate LAD profiles are packaged in `cloud2trees::leafr_for_ladderfuelsr()` which enables batch processing of individual tree point clouds and generates the input needed to extract tree CBH

```{r}
ggplot() + 
  geom_path(data = ladderfuelsr_cbh_ans$lad_profile, mapping = aes(x = lad, y = height)) +
  geom_point(data = ladderfuelsr_cbh_ans$lad_profile, mapping = aes(x = lad, y = height)) +
  # gaps data
  geom_hline(
    data = ladderfuelsr_cbh_ans$gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("gap") & !tidyselect::starts_with("gap_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "gaps")
    , linetype = "dotted"
    , lwd = 1.2
  ) + 
  # fbh data
  geom_hline(
    data = ladderfuelsr_cbh_ans$gaps_fbhs %>% 
      dplyr::select(
        tidyselect::starts_with("cbh") & !tidyselect::starts_with("cbh_")
      ) %>% 
      tidyr::pivot_longer(dplyr::everything())
    , aes(yintercept = value, color = "FBHs")
    , linetype = "dotdash"
    , lwd = 1.2
  ) +
  scale_color_manual(values = c("green4", "red"), name = "") +
  scale_y_continuous(breaks = scales::extended_breaks(10)) +
  theme_light() +
  theme(legend.position = "top")
```

note, the `gap` and `cbh` columns in `ladderfuelsr_cbh_ans$gaps_fbhs` contain the data needed for the plot ( but not the `gap_` and `cbh_` columns ;/ ).

but what are these columns?

* `cbh` - Height of the fuel layer base height (m)
* `gap` - Height of gap between fuel layers (m)

### Return CBH Metrics

```{r}
ladderfuelsr_cbh_ans$cbh_metrics %>% dplyr::glimpse()
```

what are these?

* treeID: tree ID with strings and numeric values
* treeID1: tree ID with only numeric values
* dptf: Depth of fuel layers (m) after considering distances greater than the actual height bin step
* effdist: Effective distance between consecutive fuel layers (m) after considering distances greater than any number of steps
* Hcbh: Base height of each fuel separated by a distance greater than the certain number of steps
* Hdptf: Height of the depth of fuel layers (m) after considering distances greater than the actual step
* Hdist: Height of the distance (> any number of steps) between consecutive fuel layers (m)
* Hcbh_Hdptf - Percentage of LAD values comprised in each effective fuel layer
* maxlad_Hcbh - Height of the CBH of the segmented tree based on the maximum LAD percentage
* maxlad1_Hcbh - Height of the CBH from the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_Hcbh - Height of the CBH of the segmented tree based on the maximum distance found in its profile
* last_Hcbh - Height of the CBH of the segmented tree based on the last distance found in its profile
* maxlad_ - Values of distance and fuel depth and their corresponding heights at the maximum LAD percentage
* maxlad1_ - Values of distance and fuel depth and their corresponding heights for the second fuel layer when the maximum LAD occurred in the first fuel layer but its depth <= "hdepth1_height"
* max_ - Values of distance and fuel depth and their corresponding heights at the maximum distance
* last_ - Values of distance and fuel depth and their corresponding heights at the last distance
* nlayers - Number of effective fuel layers
* max_height - Maximum height of the tree profile

there are also some plotting functions

```{r}
# Generate plots for fuels LAD metrics
plots_cbh_maxlad <- LadderFuelsR::get_plots_cbh_LAD(
  LAD_profiles = ladderfuelsr_cbh_ans$lad_profile
  , cbh_metrics = ladderfuelsr_cbh_ans$cbh_metrics
  , min_height = 0.5
)
plots_cbh_maxdist <- LadderFuelsR::get_plots_cbh_maxdist(
  LAD_profiles = ladderfuelsr_cbh_ans$lad_profile
  , cbh_metrics = ladderfuelsr_cbh_ans$cbh_metrics
  , min_height = 0.5
)
plots_cbh_lastdist <- LadderFuelsR::get_plots_cbh_lastdist(
  LAD_profiles = ladderfuelsr_cbh_ans$lad_profile
  , cbh_metrics = ladderfuelsr_cbh_ans$cbh_metrics
  , min_height = 0.5
)
# patchwork them
(plots_cbh_maxlad[[1]] + labs(title = "get_plots_cbh_LAD")) +
(plots_cbh_maxdist[[1]] + labs(title = "get_plots_cbh_maxdist")) +
(plots_cbh_lastdist[[1]] + labs(title = "get_plots_cbh_lastdist")) +
  patchwork::plot_layout(ncol = 2)
```

these plots represent the three criteria to define the CBH in a segmented tree: 

* `get_plots_cbh_LAD` = the fuel layer containing the maximum LAD percentage (column named `maxlad_Hcbh`)
* `get_plots_cbh_maxdist` = the fuel layer located at the highest distance (column named `max_Hcbh`)
* `get_plots_cbh_lastdist` = the fuel layer separated by the last effective distance (column named `last_Hcbh`)

in our [point cloud processing](#processing) we used the fuel layer separated by the last effective distance (column named `last_Hcbh`) by setting the parameter `which_cbh = "lowest"` in `cloud2trees::trees_cbh()`

```{r}
plots_cbh_lastdist[[1]]
```

### CBH on the point cloud

make a view of the CBH we estimated plotted on the point cloud

```{r}
# make a matrix to represent the cbh
x_temp <- seq(
    min(one_tree_las@data$X)
    , max(one_tree_las@data$X)
    , length.out = 2
  )
y_temp <- seq(
    min(one_tree_las@data$Y)
    , max(one_tree_las@data$Y)
    , length.out = 2
  )
xy_temp <- expand.grid(x = x_temp, y = y_temp)
z_temp <- matrix(
  rep(
    ladderfuelsr_cbh_ans$cbh_metrics$last_Hcbh
    , nrow(xy_temp)
  )
  , nrow = length(x_temp), ncol = length(y_temp)
)
# plot it
plot3D::scatter3D(
  x = one_tree_las@data$X
  , y = one_tree_las@data$Y
  , z = one_tree_las@data$Z
  , colvar = one_tree_las@data$Z
  , cex = 0.6, pch = 19
  , colkey = T
  , phi = 11
  , col = harrypotter::hp(n=50, house = "gryffindor")
  , main ="CBH shown in black"
  , surf = list(
    x = x_temp
    , y = y_temp
    , z = z_temp
    , facets = NA
    , border = "black"
    , lwd = 2
  )
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(list = ls()[grep("plots_",ls())])
remove(m, r3dDefaults, pointsByZSlice, ladderfuelsr_cbh_ans)
gc()
```

## HMD Process{#hmd}

Let's go through the process to extract height of the maximum crown diameter (HMD) from the point cloud using the same settings as in our [point cloud processing](#processing). HMD is extracted directly from the height normalized point cloud by finding the height of the non-ground point farthest from the tree center (i.e. tree top). See `cloud2trees::trees_hmd()` for a simple way to extract tree CBH from a height normalized point cloud based on a tree list with crown polygons.

### One Tree Example

this is the `tidy` data process using the data frame of the point cloud grouped by `treeID` implemented in `cloud2trees::trees_hmd()`

look at this point cloud data

```{r}
one_tree_las@data %>% dplyr::glimpse()
```

the key components we'll use are the xyz coordinates

```{r}
pts_temp <- one_tree_las@data %>% 
  dplyr::rename_with(.cols = -c(treeID), .fn = tolower) %>% 
  # get rid of ground, water, and noise points
  dplyr::filter(!classification %in% c(2,9,18)) %>% 
  dplyr::group_by(treeID) %>% # this is key
  dplyr::mutate(
    # find the highest point in the tree as the tree "center"
    is_tree_center = z==max(z)
    , tree_center_x = dplyr::first(ifelse(is_tree_center, x, NA), na_rm = T) # first in case many points are center
    , tree_center_y = dplyr::first(ifelse(is_tree_center, y, NA), na_rm = T) # first in case many points are center
    # Calculate the distance from the center to the point
    , dist_to_center = sqrt((x - tree_center_x)^2 + (y - tree_center_y)^2)
    , is_max_dist_to_center = dist_to_center == max(dist_to_center)
    # classify points for plotting
    , grp_pt = dplyr::case_when(
        is_tree_center ~ 0
        , is_max_dist_to_center ~ 1
        , T ~ 2
      ) %>% 
      factor(
        ordered = T
        , levels = 0:2
        , labels = c("tree center", "farthest point\nfrom center", "other")
      )
  ) %>% 
  dplyr::ungroup()
```

let's plot all of that in 2-D space as if we are looking down at the tree from above

```{r}
# plot
ggplot(data = pts_temp, mapping = aes(x=x,y=y,color=z, shape = grp_pt)) +
  geom_point(
    mapping = aes(
      size = ifelse(is_tree_center | is_max_dist_to_center, 1.5, 1)
    )
  ) +
  harrypotter::scale_color_hp("gryffindor", name = "Z (ht m)") +
  scale_shape_manual(values = c(17, 15, 16), name = "") +
  labs(
    title = "single tree point cloud HMD process"
    , caption = paste0(
      "HMD = "
      , round(
        pts_temp %>% 
          dplyr::filter(is_max_dist_to_center) %>% 
          dplyr::slice_head(n=1) %>% 
          dplyr::pull(z)
        , 1
      )
      , " m"
      , "\n(ht of point farthest from center)"
    )
  ) +
  theme_light() +
  theme(axis.text = element_blank()) +
  guides(size = "none")
```

look at the data on the point farthest from the tree center

```{r}
pts_temp %>% 
  dplyr::filter(is_max_dist_to_center) %>% 
  dplyr::select(x,y,z,dist_to_center,is_max_dist_to_center) %>% 
  dplyr::glimpse()
```

the HMD for this tree is contained in the `z` column

the `cloud2trees::trees_hmd()` process will give us the same answer (hopefully)

```{r, warning=FALSE}
# the function requires a LAScatalog
file_temp <- tempfile(fileext = ".las")
lidR::writeLAS(one_tree_las, file = file_temp)
# trees_hmd that
cloud2trees::trees_hmd(trees_poly = one_tree_sf, norm_las = file_temp) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(treeID, max_crown_diam_height_m)
```

success!

### Tree List Example

let's work with our clipped las data and tree crown list that we [created above](#data_ex) and use `cloud2trees::trees_hmd()` to attempt a full census of tree HMD extraction, if HMD extraction is unsuccessful we'll estimate the missing HMD values using the successfully extracted trees as training data

```{r}
# the function requires a LAScatalog
file_temp <- tempfile(fileext = ".las")
lidR::writeLAS(las, file = file_temp)
# call the function
trees_hmd_ans <- cloud2trees::trees_hmd(
  trees_poly = crowns_sf
  , norm_las = file_temp
  , tree_sample_prop = 1
  , estimate_missing_hmd = T
)
```

What is this data?

```{r}
trees_hmd_ans %>% 
  dplyr::select(treeID, tree_height_m, max_crown_diam_height_m, is_training_hmd) %>% 
  dplyr::glimpse()
```

Let’s look at the relationship between tree height and tree HMD as extracted from the point cloud. Note, that we do not expect a perfect linear relationship between tree height and HMD throughout the entire height range because HMD is also determined spatially (e.g. as a fire moves through a stand).

```{r}
trees_hmd_ans %>%
  dplyr::arrange(is_training_cbh) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = tree_height_m, y = max_crown_diam_height_m, color=is_training_hmd)
  ) + 
  ggplot2::geom_point() +
  ggplot2::labs(x = "tree ht. (m)", y = "tree HMD (m)") +
  ggplot2::scale_y_continuous(breaks = scales::extended_breaks(n=12)) +
  ggplot2::scale_x_continuous(breaks = scales::extended_breaks(n=14)) +
  ggplot2::scale_color_viridis_d(option = "turbo", begin = 0.2, alpha = 0.8, name = "is HMD\nfrom cloud?") +
  ggplot2::theme_light()
```

We can look at this data spatially too.

```{r}
trees_hmd_ans %>%
  dplyr::arrange(is_training_hmd) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(fill = max_crown_diam_height_m, color=is_training_hmd)) + 
  ggplot2::geom_sf() +
  ggplot2::scale_color_viridis_d(option = "turbo", begin = 0.2, alpha = 0.8, name = "is HMD\nfrom cloud?") +
  ggplot2::scale_fill_distiller(palette = "Greys", name = "tree HMD (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top", legend.direction = "horizontal"
    , panel.border = ggplot2::element_rect(color = "black", fill = NA)
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(lwd = 3, fill = NA))
  )
```


